---
title: "CCR_nterm_v3"
author: "AmirAli Toghani"
date: "2025-06-21"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
library(tidyverse)
library(readr)
library(Biostrings)
library(readxl)
library(ggtree)
library(bio3d)
library(ggseqlogo)
library(pheatmap)
library(scales)
library(ggridges)
library(plotly)

```

## 1. Process the taxonomy and metadata:
```{r}

# merge metadata and taxonomy
RefSeq_taxonomy_metadata <- read_xlsx("/path/to/Data_S1.xlsx)

species_meta <- RefSeq_taxonomy_metadata[,c(1,3,40,46,47)]

```

## 2. Import metadata and sequences:
```{r}
setwd("/path/to/nlrtracker")

RefSeq_files <- list.files(pattern = "_Domains.tsv", recursive = T)

RefSeq_list <- list()

# Loop through each file
for (i in seq_along(RefSeq_files)) {
  # Read the Excel file
  RefSeq_list[[i]] <- read_delim(RefSeq_files[i])
  
  # Extract the specific part of the file name and add it as a new column
  RefSeq_list[[i]] <- RefSeq_list[[i]] %>%
    mutate('Assembly Accession' = gsub(".*_(GCA|GCF)_([0-9]+\\.[0-9]+)_.*", "\\1_\\2", basename(RefSeq_files[i])))

  # Print a message for each file (optional)
  cat("Read file:", RefSeq_files[i], "and added formatted filename column\n")
}


# merge all and add the metadata
RefSeq_NLRtracker <- do.call(rbind,RefSeq_list) %>% inner_join(species_meta, by = "Assembly Accession")


RefSeq_NLRtracker <- RefSeq_NLRtracker %>% 
  mutate(ID = paste(`Order name`,`Organism Name`, seqname, Simple, sep = "_")) %>%
  # Replace spaces in the "Species" column with underscores for the "ID" column
  mutate(ID = gsub(" ", "_", ID))

```


## 3. Prepare the functions and prerequisites:
```{r}
# Function to remove duplicates based on "sequence" column
remove_duplicates_by_sequence <- function(df) {
  # Filter out rows with duplicate "sequence" values
  df %>% distinct(sequence, .keep_all = TRUE)
}

domains <- c("CNL","CNLO","CN","OCNL","CONL",
             "NL","NLO","ONL",
             "BCNL","BNL","BCN","BCCNL","BNLO","BOCNL",
             "RNL",
             "TN","TNL","TNLO","TNLJ")

```

## 4. Process the data:
```{r}
# extract the full-length NLRs
RefSeq_NLR <- filter(RefSeq_NLRtracker, RefSeq_NLRtracker$type == "CHAIN" & RefSeq_NLRtracker$Status == "NLR")

# only keep the desired domain architectures and deduplicate NLRs
RefSeq_NLR_filtered <- RefSeq_NLR[RefSeq_NLR$Simple %in% domains,]

RefSeq_NLR_filtered_deduplicated <- remove_duplicates_by_sequence(RefSeq_NLR_filtered)



# extract the NBARC domains
RefSeq_NBARC <- RefSeq_NLRtracker[RefSeq_NLRtracker$description == "NBARC",]

RefSeq_NBARC_filtered_deduplicated <- RefSeq_NBARC[RefSeq_NBARC$ID %in% RefSeq_NLR_filtered_deduplicated$ID,]


# filter out NLRs with truncated NBARC domain

RefSeq_NBARC_filtered_deduplicated_len <- filter(RefSeq_NBARC_filtered_deduplicated, 
                                                              RefSeq_NBARC_filtered_deduplicated$end - RefSeq_NBARC_filtered_deduplicated$start > 250 &
                                                              RefSeq_NBARC_filtered_deduplicated$end - RefSeq_NBARC_filtered_deduplicated$start < 400)


RefSeq_NLR_filtered_deduplicated_len <- RefSeq_NLR_filtered_deduplicated[RefSeq_NLR_filtered_deduplicated$seqname %in% RefSeq_NBARC_filtered_deduplicated_len$seqname,]

# convert the final data to biostring objects

RefSeq_NLR_filtered_deduplicated_len_seq <- AAStringSet(RefSeq_NLR_filtered_deduplicated_len$sequence)
RefSeq_NLR_filtered_deduplicated_len_seq@ranges@NAMES <- RefSeq_NLR_filtered_deduplicated_len$ID


RefSeq_NBARC_filtered_deduplicated_len_seq <- AAStringSet(RefSeq_NBARC_filtered_deduplicated_len$sequence)
RefSeq_NBARC_filtered_deduplicated_len_seq@ranges@NAMES <- RefSeq_NBARC_filtered_deduplicated_len$ID

```


## 5. Import RefPlantNLR and add it to the datasets and export:

```{r}

RefPlantNLR <- readAAStringSet("/path/to/RefPlantNLR.fasta")
RefPlantNLR_NBARC <- readAAStringSet("/path/to/RefPlantNLR_NBARC.fasta")

all_ref <- c(RefSeq_NLR_filtered_deduplicated_len_seq, RefPlantNLR)
all_ref_NBARC <- c(RefSeq_NBARC_filtered_deduplicated_len_seq, RefPlantNLR_NBARC)

writeXStringSet(all_ref_NBARC,"/path/to/all_ref_nbarc.fasta")
writeXStringSet(all_ref,"/path/to/all_ref.fasta")

# extract ZAR1 as the outgroup for smaller clades

ZAR1 <- RefPlantNLR[c("ZAR1","NbZAR1")]
ZAR1_NBARC <- RefPlantNLR_NBARC[c("ZAR1","NbZAR1")]

NRC <- all_ref[c("NRC0","NRC1","NRC2","NRC3","NRC4a","NRC4b","NRC6")]
NRC_NBARC <- all_ref_NBARC[c("NRC0","NRC1","NRC2","NRC3","NRC4a","NRC4b","NRC6")]

NRG1 <- RefPlantNLR[c("NRG1","AtNRG1.1","AtNRG1.2","AtNRG1.3")]
NRG1_NBARC <- RefPlantNLR_NBARC[c("NRG1","AtNRG1.1","AtNRG1.2")]

ADR1 <- RefPlantNLR[c("ADR1","NbADR1","ADR1-L1","ADR1-L2","ADR1-L3")]
ADR1_NBARC <- RefPlantNLR_NBARC[c("ADR1","NbADR1","ADR1-L1","ADR1-L2","ADR1-L3")]


NRG1_RPW8 <- readAAStringSet("/path/to/NRG1_RPW8.fasta")
ADR1_RPW8 <- readAAStringSet("/path/to/ADR1_RPW8.fasta")

```


## 6. Import the CCR tree and extract the CCRs:
```{r}
CCR_clade <- read.tree("~/Desktop/CCR_nterm/analysis/phylogeny/CCR_clade.tree")

CCR_meta <- RefSeq_NLR_filtered_deduplicated_len[RefSeq_NLR_filtered_deduplicated_len$ID %in% CCR_clade$tip.label,]
CCR_NBARC_meta <- RefSeq_NBARC_filtered_deduplicated_len[RefSeq_NBARC_filtered_deduplicated_len$ID %in% CCR_clade$tip.label,]

write_csv(CCR_meta,"/path/to/CCR_meta.csv")
write_csv(CCR_NBARC_meta,"/path/to/CCR_NBARC_meta.csv")


CCR_seq <- AAStringSet(CCR_meta$sequence)
CCR_seq@ranges@NAMES <- CCR_meta$ID
writeXStringSet(CCR_seq, "/path/to/CCR.fasta")

CCR_NBARC_seq <- AAStringSet(CCR_NBARC_meta$sequence)
CCR_NBARC_seq@ranges@NAMES <- CCR_NBARC_meta$ID
writeXStringSet(CCR_NBARC_seq, "/path/to/CCR_NBARC.fasta")


CCR_RPW8_seq <- subseq(CCR_seq, start = 1, end = CCR_NBARC_meta$start)
CCR_RPW8_seq_filtered <- CCR_RPW8_seq[CCR_RPW8_seq@ranges@width > 100]
writeXStringSet(CCR_RPW8_seq, "/path/to/CCR_RPW8.fasta")
writeXStringSet(CCR_RPW8_seq_filtered, "/path/to/CCR_RPW8_filtered.fasta")


# add reference sequences and export as well
c(CCR_seq, NRG1, ADR1, NRC, ZAR1) %>% writeXStringSet("/path/to/CCR_ref.fasta")

c(CCR_NBARC_seq, NRG1_NBARC, ADR1_NBARC, NRC_NBARC, ZAR1_NBARC) %>% writeXStringSet("/path/to/CCR_NBARC_ref.fasta")


```

## 7. Import data from Chia et al.

```{r}

Chia_NLRtracker <- read_xlsx("/path/to/Supplemental Data Set S1.xlsx", sheet = "4. NLR domains")

Chia_NLRtracker <- Chia_NLRtracker %>% 
  mutate(ID = paste(Lineage, Genus, Species, seqname, Simple, sep = "_")) %>%
  # Replace spaces in the "Species" column with underscores for the "ID" column
  mutate(ID = gsub(" ", "_", ID))

Chia_NLR <- filter(Chia_NLRtracker, Chia_NLRtracker$type == "CHAIN" & Chia_NLRtracker$Status == "NLR")
Chia_NLR_new <- Chia_NLR[!Chia_NLR$seqname %in% RefPlantNLR@ranges@NAMES,]
Chia_NLR_new_deduplicated <- remove_duplicates_by_sequence(Chia_NLR_new)

Chia_NBARC <- filter(Chia_NLRtracker, Chia_NLRtracker$description == "NBARC" & Chia_NLRtracker$Status == "NLR")
Chia_NBARC_new <- Chia_NBARC[!Chia_NBARC$seqname %in% RefPlantNLR_NBARC@ranges@NAMES,]
Chia_NBARC_new_duduplicated <- Chia_NBARC_new[Chia_NBARC_new$ID %in% Chia_NLR_new_deduplicated$ID,]


Chia_RNL <- Chia_NLR_new_deduplicated[Chia_NLR_new_deduplicated$Simple == "RNL",]
Chia_RNL_seq <- AAStringSet(Chia_RNL$sequence)
Chia_RNL_seq@ranges@NAMES <- Chia_RNL$ID

Chia_RNL_NBARC <- Chia_NBARC_new_duduplicated[match(Chia_RNL$ID, Chia_NBARC_new_duduplicated$ID),]
Chia_RNL_NBARC_seq <- AAStringSet(Chia_RNL_NBARC$sequence)
Chia_RNL_NBARC_seq@ranges@NAMES <- Chia_RNL_NBARC$ID


```

#### Extract the RPW8 domain:
```{r}

Chia_RPW8_seq <- subseq(Chia_RNL_seq, start = 1, end = Chia_RNL_NBARC$start)
Chia_RPW8_seq_filtered <- Chia_RPW8_seq[Chia_RPW8_seq@ranges@width > 100]
writeXStringSet(Chia_RPW8_seq, "/path/to/meme_nonflowering/Chia_RPW8.fasta")
writeXStringSet(Chia_RPW8_seq_filtered, "/path/to/meme_nonflowering/Chia_RPW8_filtered.fasta")

# merge with RefSeq RPW8 domains
RPW8_all <- c(CCR_RPW8_seq_filtered, Chia_RPW8_seq_filtered, NRG1_RPW8, ADR1_RPW8)

```

#### Export for phylogenetic analysis:
```{r}
Chia_ref_out <- c(Chia_RNL_seq, CCR_seq, NRG1, ADR1, NRC, ZAR1)
writeXStringSet(Chia_ref_out, "/path/to/phylogeny_nonflowering/Chia_ref_out.fasta")


Chia_NBARC_ref_out <- c(Chia_RNL_NBARC_seq, CCR_NBARC_seq, NRG1_NBARC, ADR1_NBARC, NRC_NBARC, ZAR1_NBARC)
writeXStringSet(Chia_NBARC_ref_out, "/path/to/phylogeny_nonflowering/Chia_NBARC_ref_out.fasta")

```

## 8. Import the subclades and extract the RPW8 domain:

```{r}

clade1 <- read.tree("/path/to/phylogeny_nonflowering/subclades/clade1.tree")
clade2 <- read.tree("/path/to/phylogeny_nonflowering/subclades/clade2.tree")
clade3 <- read.tree("/path/to/phylogeny_nonflowering/subclades/clade3.tree")
clade4 <- read.tree("/path/to/phylogeny_nonflowering/subclades/clade4.tree")

clade1_RPW8 <- RPW8_all[RPW8_all@ranges@NAMES %in% clade1$tip.label]
writeXStringSet(clade1_RPW8, "/path/to/meme_nonflowering/clade1_RPW8.fasta")

clade2_RPW8 <- RPW8_all[RPW8_all@ranges@NAMES %in% clade2$tip.label]
writeXStringSet(clade2_RPW8, "/path/to/meme_nonflowering/clade2_RPW8.fasta")

clade3_RPW8 <- RPW8_all[RPW8_all@ranges@NAMES %in% clade3$tip.label]
writeXStringSet(clade3_RPW8, "/path/to/meme_nonflowering/clade3_RPW8.fasta")

clade4_RPW8 <- RPW8_all[RPW8_all@ranges@NAMES %in% clade4$tip.label]
writeXStringSet(clade4_RPW8, "/path/to/meme_nonflowering/clade4_RPW8.fasta")


```

## 9. Import the N-term motifs and generate logo plots:
```{r}
theme_custom <- theme(
  axis.text.x = element_blank(),
  axis.text.y = element_blank(),
  axis.ticks = element_blank(),
  axis.title = element_blank(),
  legend.position = "none",
  panel.background = element_rect(fill = "transparent", colour = NA), # No fill, no border
  plot.background = element_rect(fill = "transparent", colour = NA), # No fill, no border
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  plot.margin = unit(c(0,0,0,0), "lines"),
  plot.title = element_blank(),
  plot.subtitle = element_blank(),
  plot.caption = element_blank(),
  panel.border = element_blank()
)



color_scheme <- make_col_scheme(chars = c("K","R","H",
                                          "D","E",
                                          "A","V","L","I","W","F","M","Y",
                                          "S","T","N","Q",
                                          "P","C","G"),
                                groups = c("1","1","1",
                                           "2","2",
                                           "3","3","3","3","3","3","3","3",
                                           "4","4","4","4",
                                           "5","5","5"),
                                cols = c("#FF1F1F","#FF1F1F","#FF1F1F",
                                         "#CC00CC","#CC00CC",
                                         "#3333FF","#3333FF","#3333FF","#3333FF","#3333FF","#3333FF","#3333FF","#3333FF",
                                         "#1FFF1F","#1FFF1F","#1FFF1F","#1FFF1F",
                                         "#FFAE00","#FFAE00","#FFAE00"))

theme_custom2 <- theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  legend.position = "none",
  panel.background = element_blank(),
  plot.background = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  plot.title = element_blank(),
  plot.subtitle = element_blank(),
  plot.caption = element_blank(),
  axis.line = element_line(colour = "black", size = 0.25)
)

```

```{r}
clade1_nterm_motif_seq <- readAAStringSet("/path/to/hmm_nonflowering/profiles/clade1.txt")

clade1_nterm_motif_df <- as.data.frame(clade1_nterm_motif_seq)



clade2a_nterm_motif_seq <- readAAStringSet("/path/to/hmm_nonflowering/profiles/clade2a.txt")

clade2a_nterm_motif_df <- as.data.frame(clade2a_nterm_motif_seq)



clade2b_nterm_motif_seq <- readAAStringSet("/path/to/hmm_nonflowering/profiles/clade2b.txt")

clade2b_nterm_motif_df <- as.data.frame(clade2b_nterm_motif_seq)



clade3_nterm_motif_seq <- readAAStringSet("/path/to/hmm_nonflowering/profiles/clade3.txt")

clade3_nterm_motif_df <- as.data.frame(clade3_nterm_motif_seq)



clade4_nterm_motif_seq <- readAAStringSet("/path/to/hmm_nonflowering/profiles/clade4.txt")

clade4_nterm_motif_df <- as.data.frame(clade4_nterm_motif_seq)



## Plot it

clade1_nterm_motif_p <- ggplot() +
    geom_logo(clade1_nterm_motif_df, col_scheme = color_scheme) +
    theme_light() +
    theme_custom
  
ggsave("/path/to/clade1_nterm.svg", clade1_nterm_motif_p, width = 30, height = 2)



clade2a_nterm_motif_p <- ggplot() +
    geom_logo(clade2a_nterm_motif_df, col_scheme = color_scheme) +
    theme_light() +
    theme_custom
  
ggsave("/path/to/clade2a_nterm.svg", clade2a_nterm_motif_p, width = 30, height = 2)



clade2b_nterm_motif_p <- ggplot() +
    geom_logo(clade2b_nterm_motif_df, col_scheme = color_scheme) +
    theme_light() +
    theme_custom
  
ggsave("/path/to/clade2b_nterm.svg", clade2b_nterm_motif_p, width = 24, height = 2)



clade3_nterm_motif_p <- ggplot() +
    geom_logo(clade3_nterm_motif_df, col_scheme = color_scheme) +
    theme_light() +
    theme_custom
  
ggsave("/path/to/clade3_nterm.svg", clade3_nterm_motif_p, width = 21, height = 2)



clade4_nterm_motif_p <- ggplot() +
    geom_logo(clade4_nterm_motif_df, col_scheme = color_scheme) +
    theme_light() +
    theme_custom
  
ggsave("/path/to/clade4_nterm.svg", clade4_nterm_motif_p, width = 15, height = 2)


```



## 10. Import the HMMsearch outputs:

```{r}
# clade 1
clade1_tbl <- read_table("/path/to/hmm_nonflowering/clade1_tbl.out",
                       col_names = FALSE, col_types = cols(X19 = col_skip()), 
                       skip = 3) %>% head(-10)

clade1_tbl$X5 <- as.numeric(clade1_tbl$X5)

clade1_tbl <- filter(clade1_tbl, clade1_tbl$X5 <= 0.01) # keep the significant ones
clade1_hmm <- clade1_tbl[,c(1,6)] %>% set_names(c("ID","clade1"))
clade1_hmm$clade1 <- as.numeric(clade1_hmm$clade1)

# clade 2a
clade2a_tbl <- read_table("/path/to/hmm_nonflowering/clade2a_tbl.out",
                       col_names = FALSE, col_types = cols(X19 = col_skip()), 
                       skip = 3) %>% head(-10)

clade2a_tbl$X5 <- as.numeric(clade2a_tbl$X5)

clade2a_tbl <- filter(clade2a_tbl, clade2a_tbl$X5 <= 0.01) # keep the significant ones
clade2a_hmm <- clade2a_tbl[,c(1,6)] %>% set_names(c("ID","clade2a"))
clade2a_hmm$clade2a <- as.numeric(clade2a_hmm$clade2a)

# clade 2b
clade2b_tbl <- read_table("/path/to/hmm_nonflowering/clade2b_tbl.out",
                       col_names = FALSE, col_types = cols(X19 = col_skip()), 
                       skip = 3) %>% head(-10)

clade2b_tbl$X5 <- as.numeric(clade2b_tbl$X5)

clade2b_tbl <- filter(clade2b_tbl, clade2b_tbl$X5 <= 0.01) # keep the significant ones
clade2b_hmm <- clade2b_tbl[,c(1,6)] %>% set_names(c("ID","clade2b"))
clade2b_hmm$clade2b <- as.numeric(clade2b_hmm$clade2b)

# clade 3
clade3_tbl <- read_table("/path/to/hmm_nonflowering/clade3_tbl.out",
                       col_names = FALSE, col_types = cols(X19 = col_skip()), 
                       skip = 3) %>% head(-10)

clade3_tbl$X5 <- as.numeric(clade3_tbl$X5)

clade3_tbl <- filter(clade3_tbl, clade3_tbl$X5 <= 0.01) # keep the significant ones
clade3_hmm <- clade3_tbl[,c(1,6)] %>% set_names(c("ID","clade3"))
clade3_hmm$clade3 <- as.numeric(clade3_hmm$clade3)

# clade 4
clade4_tbl <- read_table("/path/to/hmm_nonflowering/clade4_tbl.out",
                       col_names = FALSE, col_types = cols(X19 = col_skip()), 
                       skip = 3) %>% head(-10)

clade4_tbl$X5 <- as.numeric(clade4_tbl$X5)

clade4_tbl <- filter(clade4_tbl, clade4_tbl$X5 <= 0.01) # keep the significant ones
clade4_hmm <- clade4_tbl[,c(1,6)] %>% set_names(c("ID","clade4"))
clade4_hmm$clade4 <- as.numeric(clade4_hmm$clade4)

# MADA
MADA_tbl <- read_table("/path/to/hmm_nonflowering/MADA_tbl.out",
                       col_names = FALSE, col_types = cols(X19 = col_skip()), 
                       skip = 3) %>% head(-10)

MADA_tbl$X5 <- as.numeric(MADA_tbl$X5)

MADA_tbl <- filter(MADA_tbl, MADA_tbl$X5 <= 0.01) # keep the significant ones
MADA_hmm <- MADA_tbl[,c(1,6)] %>% set_names(c("ID","MADA"))
MADA_hmm$MADA <- as.numeric(MADA_hmm$MADA)

```


## 11. Import the sorted final tree and generate the hmm matrix:
```{r}
Chia_ref_out_tree <- read.tree("/path/to/hmm_nonflowering/Chia_NBARC_ref_out_trimmed_rooted_sorted.afa.tree.txt")
Chia_ref_out_IDs <- Chia_ref_out_tree$tip.label %>% as.data.frame() %>% set_names("ID")

HMM_matrix <- Reduce(function(x,y) merge(x,y, by = "ID", all = TRUE), list(Chia_ref_out_IDs, clade1_hmm, clade2a_hmm, clade2b_hmm, clade3_hmm, clade4_hmm, MADA_hmm))

HMM_matrix[is.na(HMM_matrix)] <- 0

# reorder the matrix
HMM_matrix <- HMM_matrix[match(Chia_ref_out_IDs$ID, HMM_matrix$ID),]
```

```{r}
# plot it

HMM_matrix <- HMM_matrix %>% 
  mutate(ID = factor(ID, levels = ID)) %>%
  mutate(sort = 1218:1)

HMM_matrix_melted <- HMM_matrix %>% 
  pivot_longer(cols = c(clade1, clade2a, clade2b, clade3, clade4, MADA))

HMM_matrix_melted$ID <- factor(HMM_matrix_melted$ID, levels = rev(unique(HMM_matrix_melted$ID)))

HMM_matrix_melted <- HMM_matrix_melted %>% mutate(binary = case_when(value > 0 ~ 1,
                                                                            value == 0 ~ 0))



HMM_matrix_p <- ggplot(HMM_matrix_melted, aes(x = value, y = sort)) +
  geom_segment(aes(x = 0, xend = value, y = sort, yend = sort, color = name)) +
  ylim(c(0,1218)) +
  xlim(c(0,80)) +
  facet_wrap(~ name, ncol = 6) +
  scale_color_manual(values = c("clade1" = "#616594", "clade2a" = "#999ec5", "clade2b" = "#9b89b8", "clade3" = "#efa6c9", "clade4" = "#ecc3dc")) +
  theme_linedraw(base_size = 10)

HMM_matrix_p

ggsave(plot = HMM_matrix_p, filename =  "/path/to/hmm_matrix.svg", width = 8, height = 10, dpi = "retina")

```

## 12. import the Vicia faba and Glycine sojae RNL sequences and prepare for phylogenetic analysis:
```{r}

ext_RNL <- readAAStringSet("/path/to/ext_RNL.fasta")

ext_RNL_NBARC <- readAAStringSet("/path/to/ext_RNL_NBARC.fasta")

# export for phylogenetic analysis
ext_RNL_ref_out <- c(Chia_RNL_seq, CCR_seq, ext_RNL, NRG1, ADR1, NRC, ZAR1)

ext_RNL_NBARC_ref_out <- c(Chia_RNL_NBARC_seq, CCR_NBARC_seq, ext_RNL_NBARC, NRG1_NBARC, ADR1_NBARC, NRC_NBARC, ZAR1_NBARC)
writeXStringSet(ext_RNL_NBARC_ref_out, "~/path/to/pylogeny_ext/ext_RNL_NBARC_ref_out.fasta")

```

